Scrutinizing the linear variable importance of local explanations of non-linear models with animated linear projections

### 2021-12-22 _paper.rmd
interrogate to scrutinize

---
title: "Interrogating the linear variable importance of local explanations of non-linear models with animated linear projections"
shorttitle: Cheem
keywords:
- R
- XAI
- explanable artificial intelligence
- black-box models
- local explanations
- grand tour
- manual tour
- radial tour
author:
#- name: "Nicholas Spyrison"
#     | Monash University
#     | Faculty of Information Technology
#   address: |
#     | ORCiD: 0000-0002-8417-0212
#   email: \email{nicholas.spyrison@monash.edu}
#   orcid: 0000-0002-8417-0212
#- name: "Dianne Cook"
#   affiliation: Monash University
#   address: |
#     | ORCiD: 000-0002-3813-7155
#   orcid: 000-0002-3813-7155
#- name: "Kimbal Marriott"
#   affiliation: Monash University
#   address: |
#     | ORCiD: 0000-0002-9813-0377
#   orcid: 0000-0002-9813-0377
abstract: |
 Artificial Intelligence (AI) has seen a revitalization in recent years from the use of increasingly hard-to-interpret black-box models. In such models, increased predictive power comes at the cost of opaque factor analysis, which has led to the field of eXplainable AI (XAI). XAI attempts to shed light on these models. One such approach is _local explanations_. A local explanation of a model gives a point-estimate of linear variable importance in the vicinity of one observation. We extract explanations for each observation and approximate data, and this attribution space side-by-side with linked brushing. After identifying an observation of interest, its local explanation is used as a 1D projection basis. We then manipulate the magnitude of the variable contributions with the _tour_ technique. The tour animates many projections over small changes in the projection basis as it changes the contribution of one variable. Doing so allows a user to visually explore the data space through the lens of this local explanation and test its support. The implementation of our framework is available as an __R__ package __cheem__ available on CRAN.
# documentclass: jdssv
classoption: article
output:
  bookdown::pdf_document2:
  #pdf_document:
    keep_tex: true
    toc: false
    #template: template/jdssv_template.tex
    number_sections: yes
    fig_caption: yes
bibliography: cheem_paper.bib
#appendix:
# title:
#   formatted: "Methods for understanding the variable importance of local explanations of black-box models"
#   plain:     "Methods for understanding the variable importance of local explanations of black-box models"
#   short:     "Use of the manual tour to visually interrogate local explanations"
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
 - \usepackage{setspace} ## for title page spacing
 - \usepackage{hyperref} ## for all sorts of linking
---
```{r include=FALSE, cache=FALSE}
require("knitr")
require("kableExtra")
require("magrittr")
## Work packages
require("cheem")
require("spinifex")
## chunk options
knitr::opts_chunk$set(
  fig.align  = "center",
  echo       = FALSE,
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  cache      = TRUE,
  cache.lazy = FALSE)
```

@@

<!-- Purposed approach -->
The proposed approach uses the radial, manual tour to interrogate a local explanation. After identifying an observation of interest, its explanation can be evaluated by testing the support of the structure identified by the explanation as the contributions of the variables are varied with the radial tour. We provide a free and open-source R package __cheem__ with an interactive application to facilitate analysis. We provide toy and modern datasets case studies for both classification and regression tasks.

@@

<!-- Segue -->
In summary, this figure highlights how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this instance, two players with different positions receive different profiles of variable importance to explain the prediction of their wages. In application we apply _tree SHAP_, a variant of SHAP that enjoys a lower computational complexity [@lundberg_consistent_2018]. Tree SHAP only compatible with tree-based models; we illustrate on random forests. In the following section, we will use normalized explanations as the starting projection basis to further interrogate the explanation.

>>>>

---
title: "Scrutinizing the linear variable importance of local explanations of non-linear models with animated linear projections"
shorttitle: Cheem
keywords:
- R
- XAI
- explanable artificial intelligence
- black-box models
- local explanations
- grand tour
- manual tour
- radial tour
author:
#- name: "Nicholas Spyrison"
#     | Monash University
#     | Faculty of Information Technology
#   address: |
#     | ORCiD: 0000-0002-8417-0212
#   email: \email{nicholas.spyrison@monash.edu}
#   orcid: 0000-0002-8417-0212
#- name: "Dianne Cook"
#   affiliation: Monash University
#   address: |
#     | ORCiD: 000-0002-3813-7155
#   orcid: 000-0002-3813-7155
#- name: "Kimbal Marriott"
#   affiliation: Monash University
#   address: |
#     | ORCiD: 0000-0002-9813-0377
#   orcid: 0000-0002-9813-0377
abstract: |
 Artificial Intelligence (AI) has seen a revitalization in recent years from the use of increasingly hard-to-interpret black-box models. In such models, increased predictive power comes at the cost of opaque factor analysis, which has led to the field of eXplainable AI (XAI). XAI attempts to shed light on these models. One such approach is _local explanations_. A local explanation of a model gives a point-estimate of linear variable importance in the vicinity of one observation. We extract explanations for each observation and approximate data, and this attribution space side-by-side with linked brushing. After identifying an observation of interest, its local explanation is used as a 1D projection basis. We then manipulate the magnitude of the variable contributions with the _tour_ technique. The tour animates many projections over small changes in the projection basis as it changes the contribution of one variable. Doing so allows a user to visually explore the data space through the lens of this local explanation and test its support. The implementation of our framework is available as an __R__ package __cheem__ available on CRAN.
# documentclass: jdssv
classoption: article
output:
  bookdown::pdf_document2:
  #pdf_document:
    keep_tex: true
    toc: false
    #template: template/jdssv_template.tex
    number_sections: yes
    fig_caption: yes
bibliography: cheem_paper.bib
#appendix:
# title:
#   formatted: "Methods for understanding the variable importance of local explanations of black-box models"
#   plain:     "Methods for understanding the variable importance of local explanations of black-box models"
#   short:     "cheem: radial tour scrutinizing local explanations"
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
 - \usepackage{setspace} ## for title page spacing
 - \usepackage{hyperref} ## for all sorts of linking
---
```{r include=FALSE, cache=FALSE}
require("knitr")
require("kableExtra")
require("magrittr")
## Work packages
require("cheem")
require("spinifex")
## chunk options
knitr::opts_chunk$set(
  fig.align  = "center",
  echo       = FALSE,
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  cache      = TRUE,
  cache.lazy = FALSE)
```

@@

<!-- Purposed approach -->
The proposed approach uses the radial, manual tour to scrutinize a local explanation. After identifying an observation of interest, its explanation can be evaluated by testing the support of the structure identified by the explanation as the contributions of the variables are varied with the radial tour. We provide a free and open-source R package __cheem__ with an interactive application to facilitate analysis. We provide toy and modern datasets case studies for both classification and regression tasks.

@@

<!-- Segue -->
In summary, this figure highlights how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this instance, two players with different positions receive different profiles of variable importance to explain the prediction of their wages. In application we apply _tree SHAP_, a variant of SHAP that enjoys a lower computational complexity [@lundberg_consistent_2018]. Tree SHAP only compatible with tree-based models; we illustrate on random forests. In the following section, we will use normalized explanations as the starting projection basis to further scrutinize the explanation.

### 2021-12-22 _paper.tex
interrogate to scrutinize

  pdftitle={Interrogating the linear variable importance of local explanations of non-linear models with animated linear projections},

@@

\title{Interrogating the linear variable importance of local explanations of non-linear models with animated linear projections}
\author{}
\date{\vspace{-2.5em}}

@@

The proposed approach uses the radial, manual tour to interrogate a local explanation. After identifying an observation of interest, its explanation can be evaluated by testing the support of the structure identified by the explanation as the contributions of the variables are varied with the radial tour. We provide a free and open-source R package \textbf{cheem} with an interactive application to facilitate analysis. We provide toy and modern datasets case studies for both classification and regression tasks.

@@

In summary, this figure highlights how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this instance, two players with different positions receive different profiles of variable importance to explain the prediction of their wages. In application we apply \emph{tree SHAP}, a variant of SHAP that enjoys a lower computational complexity (S. M. Lundberg, Erion, and Lee 2018). Tree SHAP only compatible with tree-based models; we illustrate on random forests. In the following section, we will use normalized explanations as the starting projection basis to further interrogate the explanation.

>>>>

  pdftitle={Scrutinizing the linear variable importance of local explanations of non-linear models with animated linear projections},

@@

\title{Scrutinizing the linear variable importance of local explanations of non-linear models with animated linear projections}
\author{}
\date{\vspace{-2.5em}}

@@

The proposed approach uses the radial, manual tour to scrutinize a local explanation. After identifying an observation of interest, its explanation can be evaluated by testing the support of the structure identified by the explanation as the contributions of the variables are varied with the radial tour. We provide a free and open-source R package \textbf{cheem} with an interactive application to facilitate analysis. We provide toy and modern datasets case studies for both classification and regression tasks.

@@

In summary, this figure highlights how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this instance, two players with different positions receive different profiles of variable importance to explain the prediction of their wages. In application we apply \emph{tree SHAP}, a variant of SHAP that enjoys a lower computational complexity (S. M. Lundberg, Erion, and Lee 2018). Tree SHAP only compatible with tree-based models; we illustrate on random forests. In the following section, we will use normalized explanations as the starting projection basis to further scrutinize the explanation.
